The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-03 RefactoringNLS/revision202307/ReviewResponses.txt

	We consider the code inside package `nlsr` to be competently coded, 
	with the advantage over `nls()` of the Levenberg-Marquardt stabilization 
	in common with `minpack.lm` and `glsnls`. The success rate is high,
        but is not perfect. However, our principal message is that we obtain 
        diagnostic information in the form of the singular values of the Jacobian 
	that allow us to try alternative starts, algorithms or programs in cases 
	where there are indications of early termination.

   We appreciate the pointers on other examples. However, a substantive comparison
   would make for a very long and likely boring paper. The BoxBOD example is, in
   many ways, similar to the Hobbs example in its nasty characteristics. 
   Some years ago, one of us created a package NISTopt to use the NIST examples 
   on function minimization tools (as per package NISTnls). See, for example, 
   https://rdrr.io/rforge/NISTopt/ 
   Experiences with these problems have led to an opinion (i.e., open to debate)
   as to the usefulness of such extreme cases. They underline a point of view that
   good diagnostics are more helpful than a "usually succeeds" level of reliability.

The section Returned results of nls() and other tools discusses the complexity 
of S3-objects of class nls, illustrated by the retrieval of the input data 
from an nls-object. It seems ill-advised to run eval(parse(text=result$data)), 
since it evaluates the name of the data object in the user’s environment, 
which may not exist. Instead, one should use model = TRUE to store the input 
data in the returned nls-object. This behavior is consistent with lm(), so 
perhaps less confusing than currently presented in the manuscript. The 
authors mention that nlxb() returns a much simpler structure of 11
items in one level, but I did not manage to find the input data object in 
the nlsr-object returned by nlxb().

Response 2:05 The section has been reworded to reflect the comments. I (JN) 
   confess to having used `lm()` only a few times in a whole career, so
   missed the parallel to that function. My work has focused on developing 
   computational tools or diagnosing extreme cases where conventional 
   statistical calculations seem to have misfired, and also do not work 
   exclusively in R.

@@

Response 2:07 Reparametrization is likely the appropriate approach if
    traditional measures of dispersion are wanted.

    Opinion: One of us (JN) has been interested for > 25 years in 
    comparing the approaches of numerical analysts and statisticians.
    The former are insterested in the nonlinear LS problem, for which
    constraints are a natural characteristic of some problems, while the
    latter want nonlinear models, where bounds constraints can be a
    nuisance to the dispersion ideas. I'll not pretend my knowledge is
    sufficiently broad to be able to capture all the nuances, but I
    believe a dialog on these matters could be very fruitful in long
    term better understanding. As far as I am aware, the literature on
    this area from a statistical viewpoint is quite thin.

    For the paper: We agree the examples are essentially artificial. They
    are intended to show how to use the features. A discussion of an upper 
    limit on the Asym parameter (i.e., weed density in some units) or the 
    scal parameter (measuring how fast this is reached) might be possible, 
    but the people who provided the problem are no longer available.

    We have added two small comments on recommended usage of bounds. One
    is to control for wild excursions in the optimization trajectory, as 
    we believe this is sensible when bounds constraints are available, 
    as in nlsr::nlxb. We also note that in cases where a simple and 
    obvious transformation is available, such as squaring a parameter to 
    ensure positivity, this is likely a better approach than imposing bounds.
    However, when there are more general bounds, transformations become more
    tricky to apply, and may de-scale a problem. We have considered the
    transfinite function used in CRAN package `adagio`, but consider that
    this needs more extensive investigation before recommending in general.

@@

5. The section Programming language makes the argument that keeping to a 
single programming language can allow for easier maintenance and upgrades, 
and that the performance penalty for using code entirely in R has been 
much reduced in recent years. I agree that there is not much of a 
performance penalty when solving small problems, such as the data examples 
included in the article. However, when solving problems using a large amount 
of data or a large number of parameters, I have experienced that the run-time 
(as well as memory usage) can become an actual bottleneck. As an example, 
consider the Penalty function I problem (Moré, Garbow, and Hillstrom (1981)) 
included in the Illustrating code section.

Response 2:08  In that we are in general agreement with this comment, we have added
    some clarification to advise that it is worthwhile being aware of how long
    calculations are taking.

    The referee's commends ant Illustrating Code led us to try the Pen1 problem
    (it is part of the funconstrain package https://github.com/jlmelville/funconstrain
    to which one of us recently added Hessian code). This showed that the R code
    of nlsr::nlfb() was sometimes faster than the compiled Fortran of minpack.lm::nls.lm.
    Moreover, the all-R code optimx::Rcgmin was much faster for larger problems --
    it does not build a matrix!

### 2023-07-04 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
mention scale=
minor updates

Using a formula specification was a principal advantage made
with `nls()` when it became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
Below we see `nls()` does poorly on the example problem. To save page space,
we use 1-line result display functions from package `nlsr`, namely `pnls()` 
and `pshort()`.

@@

```{r ex02nlsport}
unls1p<-try(nls(formula=frmu, start=stu1, data=weeddf, algorithm="port"))
pnls(unls1p)
snls1p<-try(nls(formula=frms, start=sts1, data=weeddf, algorithm="port"))
pnls(snls1p)
tnls1p<-try(nls(formula=frmt, start=stt1, data=weeddf, algorithm="port"))
pnls(tnls1p)
```

@@

```{r ex02minpack}
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
pnls(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnls(snlm1)
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
```

@@

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
pnls(ugslnls1) # to get sum of squares
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnls(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnls(tgslnls1) 
## Other algorithms
## tgslnls1a<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))
## pnls(tgslnls1a) 
## tgslnls1d<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))
## pnls(tgslnls1d) 
## tgslnls1dd<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))
## pnls(tgslnls1dd) 
## tgslnls1s2<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))
## pnls(tgslnls1s2) 
```

@@

We see that there are differences in detail, but the more important result is that
two out of three singular values are essentially 0. Our Jacobian is singular, and no 
method of the Gauss-Newton type should be able to continue. Indeed, from the parameters
reported at this saddle point, `nlsr::nlxb()` cannot proceed. We also note that
changing the start to
```
stt2<-c(Asym=1, xmid=1, scal=2)
```
allows both `nlsLM` and `gsl_nls` to find the minimum.

@@

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that `gsl_nls()` requires
a vector `y` that is the expected value of what we have called the
residual, but in this case, is actually the model. `gsl_nls` uses 
a numerical approximation for the Jacobian if the argument `jac` 
is missing. Note function `nlsr::pnlslm()` for a 1-line display of
the results of `minpack.lm::nls.lm()`.

```{r exhobbsfn, eval=TRUE}
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
pnlslm(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
pnls(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
pnls(hobgsl) # using analytic Jacobian
```

@@

```{r ex10, echo=TRUE} 
# Start MUST be feasible i.e. on or within bounds
anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
             upper=c(2,6,3), algorithm='port')
pnls(anlshob1b) #  check the answer (short form)
# nlsLM seems not to work with bounds in this example
anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls(anlsLM1b)
# also no warning if starting out of bounds, but gets a good answer!!
st4<-c(c1=4, c2=4, c3=4)
anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls(anlsLMob)
# Try nlsr::nlxb()
anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pshort(anlx1b)
```

@@

```{r ex10m, echo=TRUE}
# Hobbsmaskx.R -- masks with formula specification of the problem
require(nlsr); require(minpack.lm); traceval<-FALSE
stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
# fix first parameter
anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
print(anxbmsk1)
anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
pnls(anlM1)
anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
pnls(anlsmsk1)
# Hobbs scaled problem with bounds, formula specification
anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
                  upper=c(2,6,30))
print(anlxmsks1)
anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
             upper=c(2,6,30), algorithm='port')
pnls(anlshmsk1)
anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
                 upper=c(2,6,30))
pnls(anlsLMmsks1)

>>>>

Using a formula specification was a principal advantage made
with `nls()` when it became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
Below we see `nls()` does poorly on the example problem. To save page space,
we use 1-line result display functions from package `nlsr`, namely `pnls0()` 
and `pshort()`.

@@

```{r ex02nlsport}
unls1p<-try(nls(formula=frmu, start=stu1, data=weeddf, algorithm="port"))
pnls0(unls1p)
snls1p<-try(nls(formula=frms, start=sts1, data=weeddf, algorithm="port"))
pnls0(snls1p)
tnls1p<-try(nls(formula=frmt, start=stt1, data=weeddf, algorithm="port"))
pnls0(tnls1p)
```

@@

```{r ex02minpack}
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
pnls0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnls0(snlm1)
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
pnls0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
```

@@

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
pnls0(ugslnls1) # to get sum of squares
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnls0(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnls0(tgslnls1) 
## Other algorithms
## tgslnls1a<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))
## pnls0(tgslnls1a) 
## tgslnls1d<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))
## pnls0(tgslnls1d) 
## tgslnls1dd<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))
## pnls0(tgslnls1dd) 
## tgslnls1s2<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))
## pnls0(tgslnls1s2) 
```

@@

We see that there are differences in detail, but the more important result is that
two out of three singular values are essentially 0. Our Jacobian is singular, and no 
method of the Gauss-Newton type should be able to continue. Indeed, from the 
parameters reported at this saddle point, `nlsr::nlxb()` cannot proceed. We also 
note that changing the start to
```
stt2<-c(Asym=1, xmid=1, scal=2)
```
allows both `nlsLM` and `gsl_nls` to find the minimum. `gsl_nls` also succeeds
when the particular form of the Marquardt stabilization is changed from the
default by the control setting `scale="levenberg"`.

@@

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that `gsl_nls()` requires
a vector `y` that is the expected value of what we have called the
residual, but in this case, is actually the model. `gsl_nls` uses 
a numerical approximation for the Jacobian if the argument `jac` 
is missing. Note function `nlsr::pnls0()` for a 1-line display of
the results of `minpack.lm::nls.lm()`.

```{r exhobbsfn, eval=TRUE}
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
pnls0(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
pnls0(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
pnls0(hobgsl) # using analytic Jacobian
```

@@

```{r ex10, echo=TRUE} 
# Start MUST be feasible i.e. on or within bounds
anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
             upper=c(2,6,3), algorithm='port')
pnls0(anlshob1b) #  check the answer (short form)
# nlsLM seems not to work with bounds in this example
anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls0(anlsLM1b)
# also no warning if starting out of bounds, but gets a good answer!!
st4<-c(c1=4, c2=4, c3=4)
anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls0(anlsLMob)
# Try nlsr::nlxb()
anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pshort(anlx1b)
```

@@

```{r ex10m, echo=TRUE}
# Hobbsmaskx.R -- masks with formula specification of the problem
require(nlsr); require(minpack.lm); traceval<-FALSE
stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
# fix first parameter
anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
print(anxbmsk1)
anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
pnls0(anlM1)
anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
pnls0(anlsmsk1)
# Hobbs scaled problem with bounds, formula specification
anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
                  upper=c(2,6,30))
print(anlxmsks1)
anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
             upper=c(2,6,30), algorithm='port')
pnls0(anlshmsk1)
anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
                 upper=c(2,6,30))
pnls0(anlsLMmsks1)
